Mo' Betta Gaia Distance Measurements, \\ Using Gaia + APASS photometry

### 2017-02-09 paperDistance/paper.tex
abstract tweak
some intro text, other tweaks

%% This is emulateapj reformatting of the AASTEX sample document
%%
\documentclass[iop]{emulateapj}
\newcommand{\myemail}{landerson@simonsfoundation.org}

@@

\author{Lauren Anderson\altaffilmark{1},
        David W. Hogg\altaffilmark{1, 2, 3, 4},
        Boris Leistedt\altaffilmark{2},
        Adrian Price-Whelan\altaffilmark{5}}
\altaffiltext{1}{Center for Computational Astrophysics, Flatiron Institute}
\altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Physics, New York University}
\altaffiltext{3}{Center for Data Science, New York University}
\altaffiltext{4}{Max-Planck-Institut f\"ur Astronomie}
\altaffiltext{5}{Princeton University Observatory, Princeton University}

\begin{abstract}
The \gaia\ \tgas\ \project{Catalog} contains more than 2 million parallax measurements.
Conversion of a noisy parallax measurement into a posterior belief over distances requires inference with a prior.
Usually this prior represents some kind of belief about the Milky Way.
However, there is multi-band photometry for the \tgas\ stars from imaging surveys;
this imaging is incredibly informative about stellar distances.
Here we use color information on \tgas\ stars from \tmass\ to build a noise-deconvolved empirical prior distribution of stars in color--magnitude space.
This data-driven model contains no knowledge of the physics of stellar interiors or photospheres, nor of the Milky Way, but rather derives its precision from its generative model of noisy parallax measurements and an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm, which is an Empirical Bayes approximation to a full hierarchical model of the true parallax and photometry of every star.
The algorithm is run not in absolute-magnitude space but in a transformed space where the measurement uncertainty is closer to Gaussian.
The \xd-optimized prior is used to perform parallax and distance inferences for every star, yielding a precise stellar distance estimate and uncertainty (and full posterior) for every star.
The method is capitalizing on statistical shrinkage.
Our posterior parallax estimates are more precise than the Gaia catalog outputs by a factor of [XXX] for the median \tgas\ star; and more precise than previous examples of Bayesian distance estimates by a factor of [YYY].
We independently validate our distances by looking at members of Milky Way star clusters; for example, M67 is not visible at all in the \tgas\ parallax estimates, but appears clearly in our posterior parallax estimates.
All our results, including a posterior parallax and distance sampling for [ZZZ] \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

Given the likelihood measurements of the parallax to local stars from Gaia, and using prior information for the distances to these stars built from the data themselves, we make more precise measurements of the distances to local stars in Gaia than the parallax measurements alone. The prior, which is very similar to a color-magnitude diagram of stars, is informative because stars fall on a locus within this space. By deconvolving the set of observed points with their uncertainties we tighten this locus which therefore informs the individual measurements that they are drawn from a tighter distribution. We build this prior using Extreme Deconvolution. Extreme Deconvolution is a gaussian mixture model of data, specifically data that have heteroscedastic uncertainties, a quality that gaussin mixture models alone cannot take into account. This prior does not include any assumptions about the physical distribution of stars within the galaxy, nor does it include any stellar models.

>>>>

% This project is part of the stellarTwins project.
% Copyright 2017 the authors.

% # To-do list
% - zeroth draft the introduction
% - zeroth draft the method section
% - zeroth draft the assumptions list
% - reformat single-column

\documentclass[iop]{emulateapj}
\newcommand{\myemail}{landerson@simonsfoundation.org}

@@

\author{Lauren Anderson\altaffilmark{1},
        David W. Hogg\altaffilmark{1, 2, 3, 4},
        Boris Leistedt\altaffilmark{2},
        Adrian Price-Whelan\altaffilmark{5}}
\altaffiltext{1}{Center for Computational Astrophysics, Flatiron Institute, Simons Foundation, New York City}
\altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Physics, New York University}
\altaffiltext{3}{Center for Data Science, New York University}
\altaffiltext{4}{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
\altaffiltext{5}{Princeton University Observatory, Princeton University}

\begin{abstract}
The \gaia\ \tgas\ \project{Catalog} contains more than 2 million parallax measurements.
Conversion of a noisy parallax measurement into a posterior belief over distances requires inference with a prior.
Usually this prior represents some kind of belief about the Milky Way.
However, there is multi-band photometry for the \tgas\ stars from imaging surveys;
this imaging is incredibly informative about stellar distances.
Here we use color information on \tgas\ stars from \tmass\ to build a noise-deconvolved empirical prior distribution of stars in color--magnitude space.
This data-driven model contains no knowledge of the physics of stellar interiors or photospheres, nor of the Milky Way, but rather derives its precision from its generative model of noisy parallax measurements and an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm, which is an Empirical Bayes approximation to a full hierarchical model of the true parallax and photometry of every star.
The algorithm is run not in absolute-magnitude space but in a transformed space where the measurement uncertainty is closer to Gaussian.
The \xd-optimized prior is used to perform parallax and distance inferences for every star, yielding a precise stellar distance estimate and uncertainty (and full posterior) for every star.
Our posterior parallax estimates are more precise than the Gaia catalog outputs by a factor of [XXX] for the median \tgas\ star; and more precise than previous examples of Bayesian distance estimates by a factor of [YYY].
The precision can be attributed to the statistics concept of shrinkage.
We independently validate our distances by looking at members of Milky Way star clusters; for example, M67 is not visible at all in the \tgas\ parallax estimates, but appears clearly in our posterior parallax estimates.
All our results, including a posterior parallax and distance sampling for [ZZZ] \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

The \gaia\ Mission will deliver more than a billion stellar distances.
Only a small fraction (but large number) of these distance
measurements will be purely astrometric:
\gaia\ uses astrometric parallax to determine the distances of the closer
stars, and calibrate spectrophotometric models.
These spectrophotometric models, in turn, along with \gaia's on-board
low-resolution $B_p\,R_p$ spectrophotometry, are used to provide
distance estimates to more distant stars.
The full stack required for these distance inferences is complex.
It involves modeling not just stars, but also the dust in the Milky Way,
and the response of the telescope itself.

With projects like \project{The Cannon} (DWH CITE) and \project{Avast}
(DWH CITE), we are exploring the extent to which our predictive models of
stars could be purely data-driven or statistical.
That is, under what circumstances could the data themselves deliver
more precise or more accurate models than any theoretical or
physics-based model?

[DWH: NEEDS WORK] Given the likelihood measurements of the parallax to local stars from Gaia, and using prior information for the distances to these stars built from the data themselves, we make more precise measurements of the distances to local stars in Gaia than the parallax measurements alone. The prior, which is very similar to a color-magnitude diagram of stars, is informative because stars fall on a locus within this space. By deconvolving the set of observed points with their uncertainties we tighten this locus which therefore informs the individual measurements that they are drawn from a tighter distribution. We build this prior using Extreme Deconvolution. Extreme Deconvolution is a gaussian mixture model of data, specifically data that have heteroscedastic uncertainties, a quality that gaussin mixture models alone cannot take into account. This prior does not include any assumptions about the physical distribution of stars within the galaxy, nor does it include any stellar models.
